### 2011-10-07 tex/abstract.tex
initial commit

Explicit-state-duration hidden Markov models (EDHMM) are HMMs that have latent states consisting of both discrete state-indicator and discrete state-duration random variables.  In contrast to the implicit geometric state duration distribution possessed by the standard HMM, EDHMMs allow the direct parameterisation and estimation of per-state duration distributions. As most duration distributions have support on all positive integers, truncation or other approximations are usually required to perform EDHMM inference.  In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free inference approach for EDHMMs.  We illustrate the performance gains of this approach using synthetic data.

### 2011-10-07 tex/beam_sampling.tex
initial commit

\documentclass{article}

\title{Beam Sampling for the Explicit Duration HMM}
\author{Mike Dewar}

\begin{document}

    Problem: when sampling from an explicit duration HMM, we have to specify a maximum state duration during the forward backward algorithm, which artificially cuts off the duration distribution.

    Solution: beam sampling, as introudced by van Gael et al. Introduce an auxilliary variable $u$ such that the number of possible durations of each state is finite. For each $t$, $u_t \sim \mathrm{Uniform}(0,d_{s_t})$ where $d_{s_t}$ is the probability of the duration of the segment in state $s_t$.

### 2011-10-07 tex/discussion.tex
initial commit

We presented a beam sampler for the explicit state duration HMM. This sampler draws state sequences from the true posterior distribution without any need to make truncation approximations.
% introduction of an auxiliary variable sampling technique stochastically limits the possible transitions to a finite set, whose size is governed by the effective support of the duration distribution. Hence the computational complexity of the algorithm is governed by the support of the duration distribution instead of the magnitude of the duration itself.  
It remains future work to combine the explicit state duration HMM and the iHMM.
Python code associated with the EDHMM is available online.\footnote{http://github.com/mikedewar/EDHMM}

%Though it has not been the focus of this paper, the sampling technique could be readily extended to an infinite variant of the EDHMM, combining an infinite state representation with the infinite duration representation presented above. For more complex behaviours, the beam sampler could be applied to the SLDS to capture long behaviours without suffering a computational penalty.

### 2011-10-07 tex/edhmm.tex
initial commit

\documentclass[journal]{oldtran/IEEEtran}
\usepackage{amsmath,amssymb,graphicx,subfig}
\usepackage[usenames,dvipsnames]{color}
\usepackage{algorithm,algorithmic}
\newcommand{\mike}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}}
\newcommand{\frank}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}}
\newcommand{\simeon}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}}}
\newcommand{\E}{\textrm{E}}

\title{Inference in Hidden Markov Models with Explicit State Duration Distributions}
\author{\mike{Mike}, Chris, \frank{Frank}}

\begin{document}

\maketitle

\begin{abstract}
\input{abstract}
\end{abstract}

\section{Introduction}

\input{introduction}

    \begin{figure}[t]
        \centering
        \subfloat[][]{
            \includegraphics[width=0.25\textwidth]{EDHMM_graphical_model.pdf}
            \label{fig:graphical model}
        }
        \subfloat[][]{
            \includegraphics[width=0.25\textwidth]{EDHMM_aux_graphical_model.pdf}
         \label{fig:aux graphical model}
        }
        \caption{a) The Explicit Duration Hidden Markov Model. The time left in the current state $x_t$ is denoted $d_t$. The observation at each point in time is denoted $y_t$. b) The EDHMM with the additional auxiliary variable $u_t$ used in the beam sampler.}
    \label{fig:graphs}    
    \end{figure}
\section{Explicit Duration Hidden Markov Model}
\label{sec:Model}

\input{model}

\input{algorithm}

\section{EDHMM Inference}

\label{sec:inference}

\input{sampling}

\subsection{Related Work}

The need to accommodate explicit state duration distributions in HMMs has long been recognised. Rabiner \cite{Rabiner89} details the basic approach which expands the state space to include dwell time before applying a slightly modified Baum-Welch algorithm. This approach specifies a maximum state duration, limiting practical application 
to cases with short sequences and dwell times.
This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}. Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for implementation \emph{in-silico} \cite{Yu2006}.

Recently, a ``sticky'' variant of the hierarchical Dirichlet process HMM (HDP-HMM) has been developed \cite{Fox2008}.  The HDP-HMM has countable state-cardinality \cite{Teh06} allowing estimation of the number of states in the HMM; the sticky aspect addresses long dwell times by introducing a parameter in the prior that favours self-transition.

\section{Experiments}

\label{sec:experiments}

\input{results}

\section{Discussion}

\label{sec:dicussion}

\input{discussion}

\bibliographystyle{plain}

\bibliography{library}

\end{document}

### 2011-10-07 tex/introduction.tex
initial commit

\label{section}
Hidden Markov models (HMMs) are a fundamental tool for data analysis and exploration.  Many variants of the basic HMM have been developed in response to shortcomings in the original HMM formulation \cite{Rabiner89}.  In this paper we address inference in the explicit state duration HMM.  By state duration we mean the amount of time an HMM dwells in a state.  In the standard HMM specification, a state's duration is implicit and, a priori, distributed geometrically.

The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.  Approximate EDHMM estimation and inference can, in some cases, be performed using a modified forward-backward algorithm; specifically if either the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.   If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations shorter than the observed sequence length.  If the sequence is long then forward-backward only remains computationally tractable if only transitions between ``allowed'' durations (those that lie within pre-specified allowable intervals) are considered.   If the true state durations lie outside those intervals then the resulting model estimates will be incorrect in the sense that the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals (which may poorly describe the true duration distributions).

Our contribution is the development of a procedure for EDHMM inference that does not require any hard pre-specification of duration intervals and, as a result, is no longer approximate.  The technique we use to do this 
is borrowed from sampling procedures developed for nonparametric  Bayesian HMM variants \cite{vanGael2008}.  Our key insight is simple: the machinery developed for doing inference in HMMs with a countable number of states is precisely the same as that which is needed for doing inference in an EDHMM with duration distributions over countable support.  So, while the EDHMM is a distinctly parametric model, the tools from nonparametric Bayesian inference can be applied such that black-box inference becomes possible and, in fact, efficient.

In this work we show specifically that a ``beam-sampling'' approach  \cite{vanGael2008} works for non-approximate estimation of EDHMMs, learning both the transition structure and duration distributions simultaneously.

In demonstrating our EDHMM inference technique we consider a synthetic system  in which the state-cardinality is known and finite and the distributions over state durations are of scientific interest, we focus on the finite state, unknown duration distribution case. We show that the EDHMM beam sampler performs accurate tracking whilst capturing the duration distributions over the dwell time of each state as well as the probability of transitioning between states.

The remainder of the letter is organised as follows.  In Section~\ref{sec:Model} we introduce the EDHMM; in Section~\ref{sec:inference} we review beam-sampling for the infinite Hidden Markov Model (iHMM) \cite{Beal2002} and show how it relates to the EDHMM inference problem; and in Section~\ref{sec:experiments} we show results from using the EDHMM to model synthetic data.

### 2011-10-07 tex/model.tex
initial commit

The EDHMM captures the relationships among state $x_t$, duration $d_t$, and  observation $y_t$ over time $t$. It consists of four components: the initial state distribution, the transition distributions, the observation distributions, and the duration distributions.

We define the observation sequence $\mathcal{Y} = \{y_1, y_2, \ldots, y_T\}$; the latent state sequence $\mathcal{X} = \{x_0, x_1, x_2, \ldots, x_T\}$; and the remaining time in each segment $\mathcal{D} = \{ d_1, d_2, \ldots, d_T\}$, where $x_t \in \{ 1, 2, \ldots, K\}$ with $K$ the maximum number of states, $d_t \in \{1, 2, \ldots \}$, and $y_t \in \mathbb{R}^n$.     We assume that the Markov chain on the latent states is homogenous, i.e., that $p(x_t = i | x_{t-1}=j, A) = a_{i,j} \forall t$ where $A$ is a $K\times K$ matrix with element $a_{i,j}$ at row $i$ and column $j.$  The prior on $A$ is row-wise Dirichlet with zero prior mass on self-transitions, i.e.  $p(a_{i,:}) = \mathrm{Dir}({1}/{K-1}, \ldots, 0 , \ldots {1}/{K-1})$ where $a_{i,:}$ is a row vector and the $i$th Dirichlet parameter is $0.$  Each state is imbued with its own duration distribution $p(d_t | x_t = k) = p(d_t | \lambda_{k})$ with parameter $\lambda_k$.  Each duration distribution parameter is drawn from a prior $p(\lambda_k)$ which can be chosen in an application specific way.  The collection of all duration distribution parameters is $\lambda = \{\lambda_1, \ldots, \lambda_K\}$.  Each state is also imbued with an observation generating distribution $p(y_t  | x_t = k) = p(y_t | \theta_{k})$ with parameter $\theta_k$.  Each observation distribution parameter is 
%given 
drawn from a prior $p(\theta_k)$ also to be chosen according to the application.   The set of all observation distribution parameters is $\theta.$  In the following exposition, explicit conditional dependencies on component distribution parameters are omitted to focus on the particulars unique to the EDHMM.

In an EDHMM the transitions between states are only allowed at the end of a segment:
    \begin{equation}
        p(x_t | x_{t-1}, d_{t-1}) = 
        \begin{cases} 
            \delta(x_t, x_{t-1}) & \textrm{if $d_{t-1} > 1$} \\
            p(x_t | x_{t-1}) & \textrm{otherwise}
        \end{cases}
    \end{equation}
where the Kronecker delta $\delta(a,b) = 1$ if $a=b$ and zero otherwise. The duration distribution generates segment lengths at every state switch:
    \begin{equation}
        p(d_t | x_{t}, d_{t-1}) = 
        \begin{cases} 
            \delta(d_t, d_{t-1}-1) & \textrm{if $d_{t-1} > 1$} \\
            p(d_t | x_{t}) & \textrm{otherwise.}
        \end{cases}
    \end{equation}
The joint distribution of the EDHMM is 
\begin{multline}
    \label{eq:joint}
    p(\mathcal{X},\mathcal{D},\mathcal{Y}) =  \\ 
    p(x_0)p(d_0)\prod_{t=1}^T p(y_t | x_t, \theta) p(x_t | x_{t-1}, d_{t-1}, A) p(d_t | x_{t}, d_{t-1}, \lambda)
\end{multline}
corresponding to the graphical model in Figure \ref{fig:graphical model}. Alternative choices to define the duration variable $d_t$ exist; see \cite{Chiappa2011} for details. Algorithm \ref{alg:gen} illustrates the EDHMM as a generative model.

### 2011-10-07 tex/results.tex
initial commit

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_2_X.pdf}
        \label{fig:exp_1_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_2_Y.pdf}
        \label{fig:exp_1_data}
    }
    \caption{Example a) state and b) observation sequence generated by the explicit duration HMM. Here $K$ = 3; $p(y_t|x_t=j) = \mathrm{N}(\mu_j, 1)$ with $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$; and $p(d_t|x_t=j) = \mathrm{Poisson}(\lambda_j)$ with $\lambda_1 = 5$, $\lambda_2 = 15$, and $\lambda_3 = 20$.}
    \label{fig:experiment1_data}
\end{figure}

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_means.pdf}
        \label{fig:posterior_means}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_rates.pdf}
        \label{fig:posterior_rates}
    }
    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters.}
    \label{fig:experiment1_results}
\end{figure}
\subsection{Synthetic Data}

The first experiment uses the 500 data points shown in Figure \ref{fig:experiment1_data} generated from an EDHMM with three states. To generate the data the duration distributions were chosen to be Poisson with rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$; each observation distribution was Gaussian with means of $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$, each with a variance of 1. The transition distributions $A$ were set to
\begin{equation*}
\begin{bmatrix}
    0 & 0.3 & 0.7 \\ 0.6 & 0 & 0.4 \\ 0.3 & 0.7 & 0
\end{bmatrix}.  
\end{equation*}

For inference, broad, uninformative priors were chosen for the parameters of the duration and observation distributions. The parameters of the observation distribution for all states were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$. The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.

One thousand samples were collected from the EDHMM beam sampler after a burn-in of 500 samples. The learned posterior distribution of the state duration parameters and means of the observation distributions are shown in Figure \ref{fig:experiment1_results}.  The EDHMM achieves high accuracy in the estimated posterior distribution of the observation means, despite the overlap in observation distributions. The rate parameter distributions are reasonably estimated given the small number of observed segments.

A second experiment was performed using similar synthetic data to demonstrate the ability of the EDHMM to distinguish between states having differing duration distributions but the same observation distribution. Here the same model was used as above except data 
%was 
were
generated using $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. The same priors were used as above, and again the sampler was allowed to run for 1000 samples with a burn-in of 500 samples.

Figure~\ref{fig:experiment2_results} shows that the sampler clearly separates the high state associated with $\mu_3$ from the other two states and clearly 
%identifies the fact that there are 
reveals the presence of
two  low states with differing duration distributions.   That it is possible for the EDHMM to recover two states with differing duration distributions requires a third state.  Figure~\ref{fig:exp_2_state} shows posterior samples that indicate that the model is mixing over ambiguities about states $0$ and $1$ as it should. %As shown in Figure \ref{fig:experiment2_results}, when the state transitions to and from the long-high state to the short-low state (e.g. $t\approx 100$ and $t\approx 180$) the transitions are captured well. When the system transitions between the two low states the segmentation is less accurate, due to the inherent lack of identifiability during these periods. Instead, the space of valid state sequences is explored, each sequence obeying the duration distribution.

%Despite the loss of identifiability in the state sequence, the sampler is still able to capture reasonable estimates for the parameters of the observation and duration distributions.

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Results of the beam sampler applied to a system that generates states with identical observation distribution but differing durations. The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler after burn-in. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

Finally, Figure \ref{fig:allowed} shows that the number of allowed transitions decays with the number of iterations, significantly speeding up the runtime of the later samples after the burn-in period.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{number_transitions_visited.pdf}

\caption{Mean number of transitions visited over the first 1000 iterations of the beam sampler.}
\label{fig:allowed}
\end{figure}

### 2011-10-07 tex/sampling.tex
initial commit

Our aim is to estimate the conditional posterior distribution of the latent states ($\mathcal{X}$ and $\mathcal{D}$) and parameters ($\theta, \lambda$ and $A$) given observations $\mathcal{Y}$ by samples drawn via Markov chain Monte Carlo. Sampling $\theta$ and $A$ given $\mathcal{X}$ proceeds per usual textbook approaches \cite{Bishop06}.  Sampling $\lambda$ given $\mathcal{D}$ is straightforward in most situations.  Gibbs sampling  $\mathcal{X}$ is possible but, for reasons similar to those in the case of graphical models with chain dependency structure, one would not expect Gibbs sampling to work well for this model.  The main contribution of this paper is to show how to generate exact conditional samples of  $\mathcal{X}$ and $\mathcal{D}$ efficiently given all other random variables.

\subsection{Forward Filtering, Backward Sampling}

We can, in theory, use the forward messages from the forward backward algorithm \cite{Rabiner89} to sample the conditional posterior distribution of $\mathcal{X}$ and $\mathcal{D}.$   To do this we treat each state-duration tuple as a single random variable 
%(i.e.~$z_t = \{x_t,d_t\}$).  
(introducing the notation $z_t = \{x_t,d_t\}$).  
Doing so recovers the standard hidden Markov model structure and hence standard forward messages can be used directly.  A forward filtering, backward sampler for $\mathcal{Z} = \{z_1, \ldots, x_T\}$ conditioned on 
%everything else 
all other random variables
requires the classical forward messages:
    \begin{equation}
        \alpha_t(z_t) = 
        \sum_{z_{t-1}}
        p(z_t | z_{t-1}) 
        p(y_t|z_t) 
        \alpha_t(z_{t-1}) 
        \label{eqn:forward recursion}
    \end{equation}
   % \begin{equation}
   % \beta(z_t) = \sum_{z_{t+1}} p(z_{t+1} | z_t) p(y_{t+1}|z_{t+1}) \beta_{t+1}(z_{t+1})
   % \label{eqn:backward recursion}
    % \end{equation}
     where the transition probability can be factorised according to our modelling assumptions:
     \begin{equation}
        p(z_{t} | z_{t-1}) = p(x_t | x_{t-1}, d_{t-1}) p(d_t | d_{t-1}, x_t) p(y_t|x_t).
     \end{equation}

Unfortunately the sum in \eqref{eqn:forward recursion} has an infinite number of terms in the usual case of  duration distributions with countably infinite support. The standard approach to EDHMM inference involves truncating these sums in order to make them tractable. 
%If all durations are truncated at a maximum $d_\mathrm{max}$, yet many states are of duration far less than $d_{\mathrm{max}}$, this necessarily leads to much unnecessary computation
% if not all states have durations near $d_\mathrm{max}$. 
A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$; however, this leads to tremendous unneccesary computation for those states of duration much less than the chosen maximum.
In general, setting allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$
is highly problematic; moreover, poor choices can lead to poor inference.

%The complexity of the standard approach is always $O(T(Kd_\mathrm{max})^2)$ which is highly inefficient for segments of activity that are always long. The beam sampler below is only constrained by the effective support of the duration distribution, and hence can deal with extremely long durations with no penalty. The worst-case complexity of the beam-sampler would be achieved with a uniform duration distribution with support over $[1, d_\mathrm{max}]$.

\subsection{EDHMM Beam Sampling}

A recent contribution to inference 
in 
%is 
the infinite Hidden Markov Model (iHMM) \cite{Beal2002} suggests a way around truncation \cite{vanGael2008}.  The iHMM is an HMM with a countable number of states.  Computing the forward message for a forward filtering, backward sampler for the latent states in an iHMM also requires a sum over a countable number of elements.  
% which can be dealt with using the beam sampler \cite{vanGael2008}. Here, instead of a %potentially infinite duration-space cardinality, the iHMM faces a potentially infinite state cardinality. 
%The beam sampler combines dynamic programming with slice sampling \cite{Neal2003} in order to sample from the hierarchical Dirichlet process inherent to the iHMM.
The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice''  (see \cite{Neal2003}) auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.  The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while allowing inference to remain exact.% in the asymptotic sampling limit.

%which is used in a Gibbs sampling scheme to allow sampling from the true posterior state transition distribution.
%In the context of the EDHMM, the beam sampler allows us to sample directly from the invariant posterior $p(\mathcal{X},\mathcal{D},\mathcal{U} | \mathcal{Y})$ using standard Gibbs sampling theory. Using a block-Gibbs sampling scheme, $\mathcal{U}$ is sampled so that we are able to sample from $p(\mathcal{X},\mathcal{D} | \mathcal{U}, \mathcal{Y})$ and $p(\mathcal{U} | \mathcal{X},\mathcal{D})$. Upon convergence of the Markov chain, by simply marginalising $\mathcal{U}$ the remaining samples of state and duration are drawn directly from the posterior, without any need to choose a maximum duration.

The particular choice of auxiliary variable $u_t$ is important.  We follow  \cite{vanGael2008} in choosing $u_t$ to be conditionally distributed given the current and previous state and duration in the following way (see the graphical model in Figure \ref{fig:aux graphical model}):
\begin{equation}
    \label{eqn:slice}
    p(u_t | z_t, z_{t-1}) = 
    \frac
    {\mathbb{I}(0 < u_t < p(z_t | z_{t-1}))} 
    {p(z_t | z_{t-1})}.
\end{equation}
Given $\mathcal{U}$ it possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
%\begin{equation}
%    \hat{\alpha}_t(z_t) = p(z_t, d_t , \mathcal{Y}_1^t, \mathcal{U}_1^t)
%\end{equation}
Using notation $\mathcal{Y}_{t_1}^{t_2} = \{y_{t_1}, y_{t_1+1}, \ldots,y_{t_2}\}$  to indicate sub-ranges of a sequence, the new forward messages we compute are:
\begin{eqnarray}
   \hat{\alpha}_t(z_t) &=& 
   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t-1})   \label{eqn:scaled forward} \\
   &=& 
   \sum_{z_{t-1}}
   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t-1}) \nonumber \\
   &\propto& 
   \sum_{z_{t-1}}
   p(u_{t-1} | z_t, z_{t-1})
   p(z_t, z_{t-1} , \mathcal{Y}_1^t, \mathcal{U}_1^{t-1}) \nonumber \\
   &=& 
   \sum_{z_{t-1}}
   \mathbb{I}(0 < u_{t-1} < p(z_t | z_{t-1}))
   p(y_t|z_t) \hat{\alpha}_{t-1}(z_{t-1}) \nonumber.
\end{eqnarray}
The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$. The set of $z_t$'s for which this is true will always be finite. %\mike{PROVE THIS OR ADD A SENTENCE}
%Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:
%\begin{equation}
%    \hat{\alpha}_t(z_t) = 
%    \begin{cases} 
%            p(y_t|z_t)\sum_{z_{t-1}}\hat{\alpha}_{t-1}(z_{t-1}) & \textrm{if $u_t < p(z_t | z_{t-1})$} \\
%            0 & \textrm{otherwise}
%        \end{cases}    
%    \label{eqn:scaled forward}
%\end{equation}
%which has restricted the calculation to a finite set. 
%Care must be taken to find those transitions $p(z_t | z_{t-1})$ whose likelihood is greater than the auxiliary variable $u_t$, though this is a straightforward task for unimodal distributions.

%The auxiliary variable encodes all of the information we have about the probability of making a transition: most of the time the only valid transition will be to decrement $d_t$ and to stay in the same state. At the end of a segment the potential transitions will be filtered using the auxiliary variable, discarding unlikely transitions. The actual transition will be chosen from the remaining set, depending on the observation likelihood associated with each transition. The slightly un-intuitive aspect of this is that once this filtering has taken place, the transition probabilities are no longer used in the determination of the transition, unlike in standard forward-backward algorithms.

The backwards sampling step recursively samples a state sequence from the distribution $p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U})$
%\begin{equation}
%    p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U}).
%\end{equation}
which can expressed in terms of the forward variable:
\begin{eqnarray}
    p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U}) &\propto& p(z_{t+1},z_t, \mathcal{Y}, \mathcal{U}) \nonumber \\
    & \propto &  
        p(u_{t} | z_t, z_{t+1})p(z_{t+1}|z_t) \hat{\alpha}_t(z_t)
        %p(z_t, \mathcal{Y}_1^t, \mathcal{U}_1^t)
        %p(\mathcal{Y}_{t+1}^T, \mathcal{U}_{t+2}^T | z_{t+1}) 
        \nonumber\\
    & \propto & 
       \mathbb{I}(0 < u_{t} < p(z_{t+1} | z_t))
        \hat{\alpha}_t(z_t).
        \label{eqn:backward}
\end{eqnarray}

%Note that the indicator function used in the backwards sampler further restricts possible transitions. This restriction is conditional on the sampled $z_{t+1}$, limiting the transitions selected by in the forwards to those that could possibly result in $z_{t+1}$.

The full EDHMM beam sampler is given in Algorithm \ref{alg:beam}, which makes use of the forward recursion in \eqref{eqn:scaled forward}, the slice sampler in \eqref{eqn:slice}, and the backwards sampler in \eqref{eqn:backward}.

### 2011-10-27 tex/edhmm.tex
updated paths
small edits, mostly to do with computation complexity. One graph left...

@@
-\documentclass[journal]{oldtran/IEEEtran}
+\documentclass[journal]{../inc/oldtran/IEEEtran}
 \usepackage{amsmath,amssymb,graphicx,subfig}
@@
 \title{Inference in Hidden Markov Models with Explicit State Duration Distributions}
-\author{\mike{Mike}, Chris, \frank{Frank}}
+\author{Michael Dewar, Chris Wiggins, Frank Wood}
 
@@
         \subfloat[][]{
-            \includegraphics[width=0.25\textwidth]{EDHMM_graphical_model.pdf}
+            \includegraphics[width=0.25\textwidth]{../pic/EDHMM_graphical_model.pdf}
             \label{fig:graphical model}
@@
         \subfloat[][]{
-            \includegraphics[width=0.25\textwidth]{EDHMM_aux_graphical_model.pdf}
+            \includegraphics[width=0.25\textwidth]{../pic/EDHMM_aux_graphical_model.pdf}
          \label{fig:aux graphical model}
@@
 This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}.
-Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for implementation \emph{in-silico} \cite{Yu2006}.
+Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for practical implementation \cite{Yu2006}.
 
@@
 
-\bibliography{library}
+\bibliography{../inc/library}
 

### 2011-10-27 tex/introduction.tex
small edits, mostly to do with computation complexity. One graph left...

@@
 
-In demonstrating our EDHMM inference technique we consider a synthetic system  in which the state-cardinality is known and finite and the distributions over state durations are of scientific interest, we focus on the finite state, unknown duration distribution case.
-We show that the EDHMM beam sampler performs accurate tracking whilst capturing the duration distributions over the dwell time of each state as well as the probability of transitioning between states.
+In demonstrating our EDHMM inference technique we consider a synthetic system  in which the state-cardinality is known and finite, but where each state's duration distribution is unknown.
+We show that the EDHMM beam sampler performs accurate tracking whilst capturing the duration distributions as well as the probability of transitioning between states.
 

### 2011-10-27 tex/results.tex
updated paths
small edits, mostly to do with computation complexity. One graph left...

@@
     \subfloat[][]{
-        \includegraphics[width=0.5\textwidth]{experiment_2_X.pdf}
+        \includegraphics[width=0.5\textwidth]{../pic/experiment_2_X.pdf}
         \label{fig:exp_1_state}
@@
     \subfloat[][]{
-        \includegraphics[width=0.5\textwidth]{experiment_2_Y.pdf}
+        \includegraphics[width=0.5\textwidth]{../pic/experiment_2_Y.pdf}
         \label{fig:exp_1_data}
@@
     \subfloat[][]{
-        \includegraphics[width=0.25\textwidth]{posterior_means.pdf}
+        \includegraphics[width=0.25\textwidth]{../pic/posterior_means.pdf}
         \label{fig:posterior_means}
@@
     \subfloat[][]{
-        \includegraphics[width=0.25\textwidth]{posterior_rates.pdf}
+        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates.pdf}
         \label{fig:posterior_rates}
     }
-    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters.}
+    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters using the model described in Figure \ref{fig:exp_1_data}.}
     \label{fig:experiment1_results}
@@
     \subfloat[][]{
-        \includegraphics[width=0.5\textwidth]{experiment_3_Y.pdf}
+        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_Y.pdf}
         \label{fig:exp_2_data}
@@
     \subfloat[][]{
-        \includegraphics[width=0.5\textwidth]{experiment_3_X.pdf}
+        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_X.pdf}
         \label{fig:exp_2_state}
@@
     \subfloat[][]{
-        \includegraphics[width=0.25\textwidth]{posterior_means_exp3.pdf}
+        \includegraphics[width=0.25\textwidth]{../pic/posterior_means_exp3.pdf}
         \label{fig:posterior_means_3}
@@
     \subfloat[][]{
-        \includegraphics[width=0.25\textwidth]{posterior_rates_exp3.pdf}
+        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates_exp3.pdf}
         \label{fig:posterior_rates_3}
     }
-    \caption{Results of the beam sampler applied to a system that generates states with identical observation distribution but differing durations.
-The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler after burn-in.
+    \caption{Results of the beam sampler applied to a system that generates states with identical observation distributions but differing durations.
+The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler.
+Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$.
 Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
@@
 \begin{figure}
-    \includegraphics[width=0.5\textwidth]{number_transitions_visited.pdf}
+    \includegraphics[width=0.5\textwidth]{../pic/number_transitions_visited.pdf}
 

### 2011-10-27 tex/sampling.tex
small edits, mostly to do with computation complexity. One graph left...

@@
 % if not all states have durations near $d_\mathrm{max}$. 
-A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$; however, this leads to tremendous unneccesary computation for those states of duration much less than the chosen maximum.
-In general, setting allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$
-is highly problematic; moreover, poor choices can lead to poor inference.
+A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$, or to consider all possible durations up to the observed length of the sequence.
+This leads to either tremendous unnecessary computation, or inference that simply fails if an observed duration occurs outside the allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$.
 
-%The complexity of the standard approach is always $O(T(Kd_\mathrm{max})^2)$ which is highly inefficient for segments of activity that are always long.
+The complexity of the standard approach \cite{Yu10} is always $O(T(Kd_\mathrm{max})^2)$ which is highly inefficient for segments of activity that are always long.
 The beam sampler below is only constrained by the effective support of the duration distribution, and hence can deal with extremely long durations with no penalty.
@@
 The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice''  (see \cite{Neal2003}) auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.
-The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while allowing inference to remain exact.% in the asymptotic sampling limit.
+The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while still allowing all possible durations.% in the asymptotic sampling limit.
 
@@
 The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$.
-The set of $z_t$'s for which this is true will always be finite.
-%\mike{PROVE THIS OR ADD A SENTENCE}
+The set of $z_t$s for which this is true will always be finite and smaller than the worst case $(Kd_\mathrm{max})^2$.
 %Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:

### 2011-10-28 tex/results.tex
added a new graph to describe num transitions visited per iteration, and updated the caption. Haven't put a line in that shows the number of transitions the classical approach would have to take, as it's so large. So I just put it in the caption

@@
 
-\caption{Mean number of transitions visited over the first 1000 iterations of the beam sampler.}
+\caption{Mean number of transitions visited over the first 1000 iterations of the beam sampler after burn-in.
+Note that the standard approach would need to examine $((K-1)T)^2 = 10^6$ transitions per time point to achieve similar accuracy.}
 \label{fig:allowed}

### 2011-10-28 src/plot_num_trans_over_time.py

@@ -1,0 +1,27 @@
+import cPickle as pk
+import pylab as pb
+
+fh = open('W_exp3.dat')
+W = []
+while True:
+    try:
+        W.append(pk.load(fh))
+    except EOFError:
+        break
+
+W = pb.array(W)
+W = W.reshape((len(W)/500,500))
+
+burnin=500
+
+pb.figure(figsize=(6,2))
+pb.plot(pb.mean(W,1)[burnin:burnin+1000], 'k-', lw=1)
+pb.xlabel('iteration',{"family":"Times New Roman", "color":"black"})
+pb.ylabel('number of transitions visited',{"family":"Times New Roman", "color":"black"})
+pb.xlim((0,1000))
+pb.subplots_adjust(bottom=0.21)
+#pb.show()
+pb.savefig('../pic/number_transitions_visited.pdf')
+
+
+        

### 2011-10-29 tex/results.tex
touched stuff.  Hopefully I haven't f'ed up the length thing too much

--- 2011-10-27 tex/results.tex
@@
     }
-    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters using the model described in Figure \ref{fig:exp_1_data}.}
+    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters for the data shown in Figure \ref{fig:experiment1_data}.}
     \label{fig:experiment1_results}
@@
 
-\caption{Mean number of transitions visited over the first 1000 iterations of the beam sampler.}
+\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{MIKE}.
+Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncatation-free alternative.}
 \label{fig:allowed}

### 2011-10-29 tex/sampling.tex
touched stuff.  Hopefully I haven't f'ed up the length thing too much

@@
 Gibbs sampling  $\mathcal{X}$ is possible but, for reasons similar to those in the case of graphical models with chain dependency structure, one would not expect Gibbs sampling to work well for this model.
-The main contribution of this paper is to show how to generate exact conditional samples of  $\mathcal{X}$ and $\mathcal{D}$ efficiently given all other random variables.
+The main contribution of this paper is to show how to generate posterior samples of  $\mathcal{X}$ and $\mathcal{D}$ efficiently given all other random variables.
 
@@
 A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$, or to consider all possible durations up to the observed length of the sequence.
-This leads to either tremendous unnecessary computation, or inference that simply fails if an observed duration occurs outside the allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$.
+This leads to per sample, forward-backward computational complexity of $O(T(Kd_\mathrm{max})^2)$  and  $O(T(KT)^2)$ respectively.
+Alternatively one could risk inference that will simply fail if an actual duration lies outside hard-coded allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$.
+This yields $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$ complexity.
+The beam-sampler we propose behaves like a dynamic version of the latter, automatically scaling the allowable per-state duration intervals.
+Its use of slice-sampling auxiliary variables results in an asymptotically exact sample with no risk of incorrect inference resulting from incorrectly pre-specified duration bounds.
+We do not characterize the computational complexity of the proposed beam sampler in this work but note that it is upper bounded by $O(T(KT)^2)$ (i.e., the beam sampler admits durations of length equal to the entire sequence) but in practice is found to be as  or more efficient than the doubly hard-coded, risky approach.
 
-The complexity of the standard approach \cite{Yu10} is always $O(T(Kd_\mathrm{max})^2)$ which is highly inefficient for segments of activity that are always long.
-The beam sampler below is only constrained by the effective support of the duration distribution, and hence can deal with extremely long durations with no penalty.
-The worst-case complexity of the beam-sampler would be achieved with a uniform duration distribution with support over $[1, d_\mathrm{max}]$.
+%would be achieved with a uniform duration distribution with support over $[1, d_\mathrm{max}]$.
 
@@
 The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$.
-The set of $z_t$s for which this is true will always be finite and smaller than the worst case $(Kd_\mathrm{max})^2$.
+The beam sampler derives its computational advantage from the fact that the set of $z_t$'s for which this is true is typically small.
 %Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:

### 2011-11-02 tex/introduction.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)

@@
 The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.
-Approximate EDHMM estimation and inference can, in some cases, be performed using a modified forward-backward algorithm; specifically if either the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.
+Approximate EDHMM estimation and inference can, in some cases, be performed using a modified forward-backward algorithm; specifically if the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.
 If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations shorter than the observed sequence length.
-If the sequence is long then forward-backward only remains computationally tractable if only transitions between ``allowed'' durations (those that lie within pre-specified allowable intervals) are considered.
-If the true state durations lie outside those intervals then the resulting model estimates will be incorrect in the sense that the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals (which may poorly describe the true duration distributions).
+If the sequence is long then forward-backward only remains computationally tractable if only transitions between durations that lie within pre-specified allowable intervals are considered.
+If the true state durations lie outside those intervals then the resulting model estimates will be incorrect: the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals (which may poorly describe the true duration distributions).
 
@@
 is borrowed from sampling procedures developed for nonparametric  Bayesian HMM variants \cite{vanGael2008}.
-Our key insight is simple: the machinery developed for doing inference in HMMs with a countable number of states is precisely the same as that which is needed for doing inference in an EDHMM with duration distributions over countable support.
+Our key insight is simple: the machinery developed for inference in HMMs with a countable number of states is precisely the same as that which is needed for doing inference in an EDHMM with duration distributions over countable support.
 So, while the EDHMM is a distinctly parametric model, the tools from nonparametric Bayesian inference can be applied such that black-box inference becomes possible and, in fact, efficient.

### 2011-11-02 tex/results.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)

@@
+\subsection{Synthetic Data}
+
+The first experiment uses the 500 data points (Figure \ref{fig:experiment1_data}) generated from a three state EDHMM.
+The duration distributions were Poisson with rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$; each observation distribution was Gaussian with means of $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$, each with a variance of 1.
+The transition distributions $A$ were set to
+\begin{equation*}
+\begin{bmatrix}
+    0 & 0.3 & 0.7 \\ 0.6 & 0 & 0.4 \\ 0.3 & 0.7 & 0
+\end{bmatrix}.  
+\end{equation*}
+
 \begin{figure}
@@
 
+Broad, uninformative priors were chosen for the parameters of the duration and observation distributions.
+The the observation distribution parameters were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$.
+The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.
+
+One thousand samples were collected from the EDHMM beam sampler after a burn-in of 500 samples.
+The learned posterior distribution of the state duration parameters and means of the observation distributions are shown in Figure \ref{fig:experiment1_results}.
+The EDHMM achieves high accuracy in the estimated posterior distribution of the observation means, despite the overlap in observation distributions.
+The rate parameter distributions are reasonably estimated given the small number of observed segments.
+Figure \ref{fig:allowed} shows the mean number of transitions visited per time point over each iteration of the sampler.
+
 \begin{figure}
@@
 \end{figure}
-\subsection{Synthetic Data}
 
-The first experiment uses the 500 data points shown in Figure \ref{fig:experiment1_data} generated from an EDHMM with three states.
-To generate the data the duration distributions were chosen to be Poisson with rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$; each observation distribution was Gaussian with means of $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$, each with a variance of 1.
-The transition distributions $A$ were set to
-\begin{equation*}
-\begin{bmatrix}
-    0 & 0.3 & 0.7 \\ 0.6 & 0 & 0.4 \\ 0.3 & 0.7 & 0
-\end{bmatrix}.  
-\end{equation*}
+\begin{figure}
+    \includegraphics[width=0.5\textwidth]{../pic/number_transitions_visited.pdf}
 
-For inference, broad, uninformative priors were chosen for the parameters of the duration and observation distributions.
-The parameters of the observation distribution for all states were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$.
-The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.
+\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{fig:experiment1_results}.
+Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncation-free alternative.}
+\label{fig:allowed}
+\end{figure}
 
-One thousand samples were collected from the EDHMM beam sampler after a burn-in of 500 samples.
-The learned posterior distribution of the state duration parameters and means of the observation distributions are shown in Figure \ref{fig:experiment1_results}.
-The EDHMM achieves high accuracy in the estimated posterior distribution of the observation means, despite the overlap in observation distributions.
-The rate parameter distributions are reasonably estimated given the small number of observed segments.
+A second experiment was performed to demonstrate the ability of the EDHMM to distinguish between states having differing duration distributions but the same observation distribution.
+The same model and sampling procedure was used as above except here $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$.
 
-A second experiment was performed using similar synthetic data to demonstrate the ability of the EDHMM to distinguish between states having differing duration distributions but the same observation distribution.
-Here the same model was used as above except data
-%was 
-were
-generated using $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. The same priors were used as above, and again the sampler was allowed to run for 1000 samples with a burn-in of 500 samples.
-
-Figure~\ref{fig:experiment2_results} shows that the sampler clearly separates the high state associated with $\mu_3$ from the other two states and clearly 
+Figure~\ref{fig:experiment2_results} shows that the sampler clearly separates the high state associated with $\mu_3$ from the other states and clearly 
 %identifies the fact that there are 
@@
 two  low states with differing duration distributions.   
-That it is possible for the EDHMM to recover two states with differing duration distributions requires a third state.
+%That it is possible for the EDHMM to recover two states with differing duration distributions requires a third state.  
 Figure~\ref{fig:exp_2_state} shows posterior samples that indicate that the model is mixing over ambiguities about states $0$ and $1$ as it should.
@@
     }
-    \caption{Results of the beam sampler applied to a system that generates states with identical observation distributions but differing durations.
-The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler.
+    \caption{Beam sampler results applied to a system with identical observation distributions but differing durations.
+Observations are shown in a); true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler.
 Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$.
@@
 \end{figure}
-
-Finally, Figure \ref{fig:allowed} shows that the number of allowed transitions decays with the number of iterations, significantly speeding up the runtime of the later samples after the burn-in period.
-
-\begin{figure}
-    \includegraphics[width=0.5\textwidth]{../pic/number_transitions_visited.pdf}
-
-\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{MIKE}.
-Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncatation-free alternative.}
-\label{fig:allowed}
-\end{figure}

### 2011-11-02 tex/sampling.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)

@@
 %The beam sampler combines dynamic programming with slice sampling \cite{Neal2003} in order to sample from the hierarchical Dirichlet process inherent to the iHMM.
-The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice''  (see \cite{Neal2003}) auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.
+The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice'' \cite{Neal2003} auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.
 The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while still allowing all possible durations.% in the asymptotic sampling limit.

### 2011-11-02 tex/introduction.tex
my final (I hope) edits

@@
 The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.
-Approximate EDHMM estimation and inference can, in some cases, be performed using a modified forward-backward algorithm; specifically if the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.
-If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations shorter than the observed sequence length.
+EDHMM estimation and inference can be performed using the forward-backward algorithm; specifically if the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.
+If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations up to the observed sequence length.
 If the sequence is long then forward-backward only remains computationally tractable if only transitions between durations that lie within pre-specified allowable intervals are considered.
-If the true state durations lie outside those intervals then the resulting model estimates will be incorrect: the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals (which may poorly describe the true duration distributions).
+If the true state durations lie outside those intervals then the resulting model estimates will be incorrect: the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals.
+% (which may poorly describe the true duration distributions).
 
@@
 Our key insight is simple: the machinery developed for inference in HMMs with a countable number of states is precisely the same as that which is needed for doing inference in an EDHMM with duration distributions over countable support.
-So, while the EDHMM is a distinctly parametric model, the tools from nonparametric Bayesian inference can be applied such that black-box inference becomes possible and, in fact, efficient.
+So, while the EDHMM is a distinctly parametric model, the tools from nonparametric Bayesian inference can be applied such that black-box inference becomes possible and, in practice, efficient.
 

### 2011-11-02 tex/introduction.tex
a little more nudging of the introduction text

@@
 
-Our contribution is the development of a procedure for EDHMM inference that does not require any hard pre-specification of duration intervals and, as a result, is no longer approximate.
+Our contribution is the development of a procedure for EDHMM inference that does not require any hard pre-specification of duration intervals, is efficient in practice, and, as it is an asymptotically exact procedure, does not risk incorrect inference.
 The technique we use to do this
@@
 
-In this work we show specifically that a ``beam-sampling'' approach  \cite{vanGael2008} works for non-approximate estimation of EDHMMs, learning both the transition structure and duration distributions simultaneously.
-
+In this work we show specifically that a ``beam-sampling'' approach  \cite{vanGael2008} works for  estimating EDHMMs, learning both the transition structure and duration distributions simultaneously.
 In demonstrating our EDHMM inference technique we consider a synthetic system  in which the state-cardinality is known and finite, but where each state's duration distribution is unknown.

### 2011-11-03 tex/edhmm.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)
tiny adjustments to the size of the graphical models. I think they're as big as they can get without falling into the gutter
final update

@@
 \usepackage{algorithm,algorithmic}
-\newcommand{\mike}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}}
-\newcommand{\frank}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}}
-\newcommand{\simeon}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}}}
 \newcommand{\E}{\textrm{E}}
 
+\begin{document}
+
 \title{Inference in Hidden Markov Models with Explicit State Duration Distributions}
-\author{Michael Dewar, Chris Wiggins, Frank Wood}
+\author{Michael Dewar, Chris Wiggins, Frank Wood}%
 
-\begin{document}
+%\thanks{Michael Dewar is with bitly, New York City}%
+%\thanks{Chris Wiggins is with the Department of Applied Physics and Applied Mathematics, Columbia University}%
+%\thanks{Frank Wood is the the Department of Statistics, Columbia Univeristy}%
 
@@
         \subfloat[][]{
-            \includegraphics[width=0.25\textwidth]{../pic/EDHMM_graphical_model.pdf}
+            \includegraphics[width=0.23\textwidth]{../pic/EDHMM_graphical_model.pdf}
             \label{fig:graphical model}
@@
         \subfloat[][]{
-            \includegraphics[width=0.25\textwidth]{../pic/EDHMM_aux_graphical_model.pdf}
+            \includegraphics[width=0.23\textwidth]{../pic/EDHMM_aux_graphical_model.pdf}
          \label{fig:aux graphical model}

### 2011-12-17 tex/edhmm.tex
made reviewer typo edits

@@
 This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}.
-Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for practical implementation \cite{Yu2006}.
+Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96}, speech synthesis \cite{Zen07}, and evolved into symmetric approaches suitable for practical implementation \cite{Yu2006}.
 

### 2011-12-17 tex/introduction.tex
made reviewer typo edits

@@
 Many variants of the basic HMM have been developed in response to shortcomings in the original HMM formulation \cite{Rabiner89}.
-In this paper we address inference in the explicit state duration HMM.
+In this paper we address inference in the explicit state duration HMM (EDHMM).
 By state duration we mean the amount of time an HMM dwells in a state.
@@
 
-The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.
-EDHMM estimation and inference can be performed using the forward-backward algorithm; specifically if the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.
-If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations up to the observed sequence length.
-If the sequence is long then forward-backward only remains computationally tractable if only transitions between durations that lie within pre-specified allowable intervals are considered.
-If the true state durations lie outside those intervals then the resulting model estimates will be incorrect: the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals.
-% (which may poorly describe the true duration distributions).
+The explicit state duration HMM (EDHMM) \cite{Rabiner89}, a member of the Hidden Semi Markov Models \cite{Yu10}, was developed to 
+allow explicit parameterization and direct inference of state duration 
+distributions.   EDHMM estimation and inference can be performed using the 
+forward-backward algorithm; specifically if the sequence is short or a tight 
+``allowable'' duration interval for each state is hard-coded a priori 
+\cite{Yu2006}.   If the sequence is short then forward-backward can be run 
+on a state representation that allows for all possible durations up to the 
+observed sequence length.  If the sequence is long then forward-backward 
+only remains computationally tractable if only transitions between durations 
+that lie within pre-specified allowable intervals are considered.   If the 
+true state durations lie outside those intervals then the resulting model 
+estimates will be incorrect: the learned duration distributions will reflect 
+only what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).
 

### 2011-12-17 tex/results.tex
final update
made reviewer typo edits

@@
 Broad, uninformative priors were chosen for the parameters of the duration and observation distributions.
-The the observation distribution parameters were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$.
+The observation distribution parameters were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$.
 The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.
@@
 Observations are shown in a); true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler.
-Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$.
+Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$ and the duration distributions have rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$.
 Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}

### 2011-12-17 tex/sampling.tex
my final (I hope) edits
made reviewer typo edits

@@
 Doing so recovers the standard hidden Markov model structure and hence standard forward messages can be used directly.
-A forward filtering, backward sampler for $\mathcal{Z} = \{z_1, \ldots, x_T\}$ conditioned on
+A forward filtering, backward sampler for $\mathcal{Z} = \{z_1, \ldots, z_T\}$ conditioned on
 %everything else 
@@
      \begin{equation}
-        p(z_{t} | z_{t-1}) = p(x_t | x_{t-1}, d_{t-1}) p(d_t | d_{t-1}, x_t) p(y_t|x_t).
+        p(z_{t} | z_{t-1}) = p(x_t | x_{t-1}, d_{t-1}) p(d_t | d_{t-1}, x_t).
      \end{equation}
@@
 \end{equation}
-Given $\mathcal{U}$ it possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
+where $\mathbb{I}(\cdot)$ returns one if its operand is true and zero otherwise. Given $\mathcal{U}$ it is possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
 %\begin{equation}
@@
    \hat{\alpha}_t(z_t) &=& 
-   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t-1})   \label{eqn:scaled forward} \\
+   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t})   \label{eqn:scaled forward} \\
    &=& 
    \sum_{z_{t-1}}
-   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t-1}) \nonumber \\
+   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t}) \nonumber \\
    &\propto& 
    \sum_{z_{t-1}}
-   p(u_{t-1} | z_t, z_{t-1})
+   p(u_{t} | z_t, z_{t-1})
    p(z_t, z_{t-1} , \mathcal{Y}_1^t, \mathcal{U}_1^{t-1}) \nonumber \\
@@
    \sum_{z_{t-1}}
-   \mathbb{I}(0 < u_{t-1} < p(z_t | z_{t-1}))
+   \mathbb{I}(0 < u_{t} < p(z_t | z_{t-1}))
    p(y_t|z_t) \hat{\alpha}_{t-1}(z_{t-1}) \nonumber.
 \end{eqnarray}
-The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$.
+The indicator function $\mathbb{I}$ results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$.
 The beam sampler derives its computational advantage from the fact that the set of $z_t$'s for which this is true is typically small.
@@
 
-The backwards sampling step recursively samples a state sequence from the distribution $p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U})$
+The backwards sampling step recursively samples a state sequence from the distribution $p(z_{t-1} | z_{t}, \mathcal{Y}, \mathcal{U})$
 %\begin{equation}
@@
 \begin{eqnarray}
-    p(z_t | z_{t+1}, \mathcal{Y}, \mathcal{U}) &\propto& p(z_{t+1},z_t, \mathcal{Y}, \mathcal{U}) \nonumber \\
+    p(z_{t-1} | z_{t}, \mathcal{Y}, \mathcal{U}) &\propto& p(z_{t},z_{t-1}, \mathcal{Y}, \mathcal{U})  \label{eqn:backward} \\
     & \propto &  
-        p(u_{t} | z_t, z_{t+1})p(z_{t+1}|z_t) \hat{\alpha}_t(z_t)
+        p(u_{t} | z_t, z_{t-1})p(z_{t}|z_{t-1}) \hat{\alpha}_{t-1}(z_{t-1})
         %p(z_t, \mathcal{Y}_1^t, \mathcal{U}_1^t)
@@
     & \propto & 
-       \mathbb{I}(0 < u_{t} < p(z_{t+1} | z_t))
-        \hat{\alpha}_t(z_t).
-        \label{eqn:backward}
+       \mathbb{I}(0 < u_{t} < p(z_{t} | z_{t-1}))
+        \hat{\alpha}_{t-1}(z_{t-1}).\nonumber
 \end{eqnarray}

### 2011-12-20 tex/abstract.tex
my final (I hope) edits
Frank's last (hopefully) changes

@@
 In contrast to the implicit geometric state duration distribution possessed by the standard HMM, EDHMMs allow the direct parameterisation and estimation of per-state duration distributions.
-As most duration distributions have support on all positive integers, truncation or other approximations are usually required to perform EDHMM inference.
-In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free inference approach for EDHMMs.
-We illustrate the performance gains of this approach using synthetic data.
+As most duration distributions are defined over the positive integers, truncation or other approximations are usually required to perform EDHMM inference.
+In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free, black-box inference procedure for EDHMMs.
+%We illustrate the performance gains of this approach using synthetic data.

### 2011-12-20 tex/introduction.tex
Frank's last (hopefully) changes

@@
 
-The explicit state duration HMM (EDHMM) \cite{Rabiner89}, a member of the Hidden Semi Markov Models \cite{Yu10}, was developed to 
+The EDHMM  (or, equivalently, the hidden semi-Markov model \cite{Yu10}) was developed to 
 allow explicit parameterization and direct inference of state duration 
 distributions.   EDHMM estimation and inference can be performed using the 
-forward-backward algorithm; specifically if the sequence is short or a tight 
+forward-backward algorithm; though only if the sequence is short or a tight 
 ``allowable'' duration interval for each state is hard-coded a priori 
@@
 true state durations lie outside those intervals then the resulting model 
-estimates will be incorrect: the learned duration distributions will reflect 
-only what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).
+estimates will be incorrect: the learned duration distributions can only reflect 
+ what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).
 

### 2011-12-20 tex/results.tex
Frank's last (hopefully) changes

@@
 
-\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{fig:experiment1_results}.
-Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncation-free alternative.}
+\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in sweeps on data from Figure \ref{fig:experiment1_results}.
+Consider this in comparison to the $(KT)^2 = O(10^6)$ per time point transitions that would need to be considered by standard forward backward without truncation, a surely-safe, truncation-free, but computationally impractical alternative.}
 \label{fig:allowed}

### 2011-12-20 tex/sampling.tex
Frank's last (hopefully) changes

@@
 Sampling $\lambda$ given $\mathcal{D}$ is straightforward in most situations.
-Gibbs sampling  $\mathcal{X}$ is possible but, for reasons similar to those in the case of graphical models with chain dependency structure, one would not expect Gibbs sampling to work well for this model.
-The main contribution of this paper is to show how to generate posterior samples of  $\mathcal{X}$ and $\mathcal{D}$ efficiently given all other random variables.
+Indirect Gibbs sampling  of $\mathcal{X}$ is possible using auxiliary state-change indicator variables, but for reasons similar to those in \cite{Goldwater2009}, such a sampler will not mix well.
+The main contribution of this paper is to show how to generate posterior samples of  $\mathcal{X}$ and $\mathcal{D}$.
 
@@
 
-Unfortunately the sum in \eqref{eqn:forward recursion} has an infinite number of terms in the usual case of  duration distributions with countably infinite support.
-The standard approach to EDHMM inference involves truncating these sums in order to make them tractable.
-%If all durations are truncated at a maximum $d_\mathrm{max}$, yet many states are of duration far less than $d_{\mathrm{max}}$, this necessarily leads to much unnecessary computation
-% if not all states have durations near $d_\mathrm{max}$. 
-A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$, or to consider all possible durations up to the observed length of the sequence.
-This leads to per sample, forward-backward computational complexity of $O(T(Kd_\mathrm{max})^2)$  and  $O(T(KT)^2)$ respectively.
-Alternatively one could risk inference that will simply fail if an actual duration lies outside hard-coded allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$.
-This yields $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$ complexity.
-The beam-sampler we propose behaves like a dynamic version of the latter, automatically scaling the allowable per-state duration intervals.
-Its use of slice-sampling auxiliary variables results in an asymptotically exact sample with no risk of incorrect inference resulting from incorrectly pre-specified duration bounds.
-We do not characterize the computational complexity of the proposed beam sampler in this work but note that it is upper bounded by $O(T(KT)^2)$ (i.e., the beam sampler admits durations of length equal to the entire sequence) but in practice is found to be as  or more efficient than the doubly hard-coded, risky approach.
+Unfortunately the sum in \eqref{eqn:forward recursion} has at worst an infinite number of terms in the case of  duration distributions with countably infinite support and at best a very large number of terms in the case of long sequences.
+The standard approach to EDHMM inference involves truncating considered durations to only those that lie between $d_\mathrm{min}$ and $d_\mathrm{max}$ or computation involving all possible durations up to the observed length of the sequence ($d_\mathrm{min}=0, d_\mathrm{max}=T$).
+This leads to per-sample, forward-backward computational complexity of $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$.
+Truncation yields inference that will simply fail if an actual duration lies outside hard-coded allowable duration intervals.
+Considering all possible durations up to length $T$ is often computationally impossible.
+%This yields $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$ complexity.
+The beam-sampler we propose behaves like a dynamic version of the truncation approach, automatically defining and scaling per-state duration truncation intervals.
+Better though, the way it does this results in an asymptotically exact sample with no risk of incorrect inference resulting from incorrectly pre-specified duration truncations.
+We do not characterize the computational complexity of the proposed beam sampler in this work but note that it is upper bounded by $O(T(KT)^2)$ (i.e., the beam sampler admits durations of length equal to the entire sequence) but in practice is found to be as  or more efficient than the risky hard-truncation approach.
 

### 2011-12-21 tex/results.tex
compressed the paper back down to 4 pages

@@
 \begin{figure}
-    \includegraphics[width=0.5\textwidth]{../pic/number_transitions_visited.pdf}
+    \includegraphics[width=0.45\textwidth]{../pic/number_transitions_visited.pdf}
 
@@
 The same model and sampling procedure was used as above except here $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$.
-
 Figure~\ref{fig:experiment2_results} shows that the sampler clearly separates the high state associated with $\mu_3$ from the other states and clearly 
@@
 two  low states with differing duration distributions.
-%That it is possible for the EDHMM to recover two states with differing duration distributions requires a third state.  
 Figure~\ref{fig:exp_2_state} shows posterior samples that indicate that the model is mixing over ambiguities about states $0$ and $1$ as it should.
@@
     }
-    \caption{Beam sampler results applied to a system with identical observation distributions but differing durations.
-Observations are shown in a); true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler.
-Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$ and the duration distributions have rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$.
-Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
+    \caption{Beam sampler results from a system with identical observation distributions but differing durations.
+Observations are shown in a); true states in b) overlaid with 20 state traces produced by the sampler.
+Here we have parameters $\mu_1 = \mu_2 = 0$, $\mu_3 = 3$ and $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$.
+Samples from the posterior observation-mean and duration-rate distributions are shown in c) and d), respectively.}
     \label{fig:experiment2_results}

### 2011-12-21 src/plot_num_trans_over_time.py

@@ -14,7 +14,7 @@ W = W.reshape((len(W)/500,500))
 
 burnin=500
 
-pb.figure(figsize=(6,2))
+pb.figure(figsize=(6.5,2))
 pb.plot(pb.mean(W,1)[burnin:burnin+1000], 'k-', lw=1)
 pb.xlabel('iteration',{"family":"Times New Roman", "color":"black"})
 pb.ylabel('number of transitions visited',{"family":"Times New Roman", "color":"black"})

### 2011-12-23 tex/response.tex
added the first draft of the response to reviwers

\documentclass{article}
\usepackage{amsmath,amssymb,graphicx,subfig}

\title{SPL-10473-2011, Inference in Hidden Markov Models with Explicit State Duration Distributions - Response to Reviewers}

\begin{document}
\maketitle
We would like to thank the reiviwers for their encouraging comments. Below we respond to each reviwer in turn.

\section*{Reviwer 1}

We would like to thank reviewer 1 for the detailed response. We have corrected all the typos mentioned and included the reference on speech synthesis.

Reviwer 1 has suggested the comparison of our proposed method with the Forward Backward algorithm and Gibbs sampling. We do not have additional room to introduce a comparison to the letter, and neither of these comparisons are straightforward.

The first suggested comparison is Gibbs sampler, which will not mix well and would be computationally intractable to apply to the first example in our letter. We have edited the introduction to section 3 and included an additional reference to support this.

The second suggestion was to use the forward backward algorithm to perform inference. This can be achieved by artificially setting the auxilliary variable $u$ to zero, and hence accepting each transition in the sampler. This introduces a new problem: how to choose a range of possible durations over which to sample. To perform this in a manner that is guaranteed not to break, we would have to allow any duration up to the length of the sequence. This is computationally infeasable (see the caption to Figure 4), but it is possible to examine the likelihood of the model as we increase $d_\mathrm{max}$. We have placed the code for this experiment in the code repository and include a quick graph below. This graph shows the mean likelihood of the samples after burn-in as the maximum duration increases, and shows how the basic algorithm suffers without knowledge of the duration distribuion. The mean likelihood of the beam sampler is shown in purple for comparison.

\begin{figure}
	\includegraphics[width=0.8\textwidth]{../pic/likelihood_over_dmax.pdf}
	\caption{Mean likelihood over samples after burn-in as the range of possible durations increases (blue line). Shown for comparison is the mean likelihood of the beam sampler (purple line)}
\end{figure}

\section*{Reviwewr 2}

We thank the reviwer for their commenets, we have adressed the typos.

\section*{Reviewer 3}

Reviwer 3 has highlighted two omissions in the paper. The first was that we did not make clear the fact that the Explicit Duration HMM is a type of Hidden Semi Markov Model, which is now made clear in the second paragraph of the introduction, along with an additional citation to an HSMM review detailing the relationship. The second was that we did not report on the results of the transition rate parameter estimates. While we do not report any detailed results for these parameters, as they are not the focus of the paper, we have made sure to report that this estimation is done. The reader is able to reproduce our experiments using the code online for more detailed investigation if necessary.

We were unable to include all of the references mentioned due to the space limitations of the paper, however we have made sure to include one speech-synthesis paper, also suggested by revier 1, by Zen et al to make sure this application area is represented in the text.

\section*{Reviwer 4}

We thank the reviewer for their comments, and have fixed the typos. 
\end{document}

### 2011-12-23 tex/response.tex
made changes to the response

@@
 \maketitle
-We would like to thank the reiviwers for their encouraging comments. Below we respond to each reviwer in turn.
+We would like to thank the reviewers for their encouraging comments. Below we respond to each reviewer in turn.
 
@@
 
-Reviwer 1 has suggested the comparison of our proposed method with the Forward Backward algorithm and Gibbs sampling.
-We do not have additional room to introduce a comparison to the letter, and neither of these comparisons are straightforward.
+Reviewer 1 has suggested the comparison of our proposed method with Gibbs sampling and the Forward Backward algorithm.
+We found that we do not have sufficient space to introduce a meaningful comparison to the either.
+We do here, however, show additional experiments which may help address the reviewer's concern.
+We hope that the corrections we made to our prose will be sufficient to explain away many of the reviewers' concerns about the note.
 
-The first suggested comparison is Gibbs sampler, which will not mix well and would be computationally intractable to apply to the first example in our letter.
-We have edited the introduction to section 3 and included an additional reference to support this.
+The first suggested comparison is to a Gibbs sampler.
+The Gibbs sampler was too casually mentioned in the original text as Gibbs sampling would require block-sampling in a model augmented with state-change-boundary auxiliary indicator variables.
+We have edited the introduction to section 3 and included an additional reference to a paper that samples in this representation and reports the kind of mixing/computational problems that we previously hinted at.
 
-The second suggestion was to use the forward backward algorithm to perform inference.
-This can be achieved by artificially setting the auxilliary variable $u$ to zero, and hence accepting each transition in the sampler.
-This introduces a new problem: how to choose a range of possible durations over which to sample.
-To perform this in a manner that is guaranteed not to break, we would have to allow any duration up to the length of the sequence.
-This is computationally infeasable (see the caption to Figure 4), but it is possible to examine the likelihood of the model as we increase $d_\mathrm{max}$.
+The second suggestion was to use the forward-backward algorithm to perform inference.
+This can be achieved by artificially setting the auxiliary variable $u$ to zero, and hence accepting each transition in the sampler, but then requires $O(T^3K^2)$ total cost which is impractical for even the short toy data examples provided in the paper.
+To get out of this re-introduces the problem we set out to solve: how to choose a range of possible durations over which to sample.
+One way to do this in a manner that is guaranteed not to break, is to allow any duration up to the length of the sequence.
+Since this is computationally infeasable (see the caption to Figure 4), we can, instead, show the achieved expected joint-likelihood of the model and the data as we increase $d_\mathrm{max}$.
+This unsurprisingly breaks exactly as we would expect it to.
+This graph shows the mean likelihood of the samples after burn-in as the maximum duration increases, and shows one way that the truncated forward/backward algorithm suffers when the true durations are not in the support of the truncation windows.
 We have placed the code for this experiment in the code repository and include a quick graph below.
-This graph shows the mean likelihood of the samples after burn-in as the maximum duration increases, and shows how the basic algorithm suffers without knowledge of the duration distribuion.
 The mean likelihood of the beam sampler is shown in purple for comparison.
+
+MIKE: IF THERE ARE NO STATES OF DURATIONS GREATER THAN 25 IN THE DATA THEN I DONT BELIEVE THIS FIGURE.
+IF $d_{max}$ IS SET LARGER THAN THE BIGGEST OBSERVED SEQUENCE THEN THESE FIGURES SHOULD OVERLAP.
+I DON'T KNOW WHAT DATA THIS FIGURE CORRESPONDS TO AND WHAT DURATIONS ARE IN THE DATA.
+IT MIGHT MAKE SENSE TO RUN THESE EXPERIMENTS OUT TO OVERLAP.
+IT SHOULD BE POSSIBLE.
 
@@
 
-We thank the reviwer for their commenets, we have adressed the typos.
+We thank the reviewer for their comments, we have addressed the typos.
 
@@
 
-Reviwer 3 has highlighted two omissions in the paper.
+Reviewer 3 has highlighted two omissions in the paper.
 The first was that we did not make clear the fact that the Explicit Duration HMM is a type of Hidden Semi Markov Model, which is now made clear in the second paragraph of the introduction, along with an additional citation to an HSMM review detailing the relationship.
@@
 
-We were unable to include all of the references mentioned due to the space limitations of the paper, however we have made sure to include one speech-synthesis paper, also suggested by revier 1, by Zen et al to make sure this application area is represented in the text.
+We were unable to include all of the references mentioned due to the space limitations of the paper, however we have made sure to include one speech-synthesis paper, also suggested by reviewer 1, by Zen et al to make sure this application area is represented in the text.
 
-\section*{Reviwer 4}
+\section*{Reviewer 4}
 

### 2012-01-14 tex/response.tex
final edits

@@
 \title{SPL-10473-2011, Inference in Hidden Markov Models with Explicit State Duration Distributions - Response to Reviewers}
+
+\author{Mike Dewar, Chris Wiggins, and Frank Wood}
 
@@
 Reviewer 1 has suggested the comparison of our proposed method with Gibbs sampling and the Forward Backward algorithm.
-We found that we do not have sufficient space to introduce a meaningful comparison to the either.
+We found that we do not have sufficient space to introduce a meaningful comparison to either.
 We do here, however, show additional experiments which may help address the reviewer's concern.
-We hope that the corrections we made to our prose will be sufficient to explain away many of the reviewers' concerns about the note.
+We hope that the corrections we made to our prose will be sufficient to explain away many of the reviewer's concerns about the letter.
 
@@
 The second suggestion was to use the forward-backward algorithm to perform inference.
-This can be achieved by artificially setting the auxiliary variable $u$ to zero, and hence accepting each transition in the sampler, but then requires $O(T^3K^2)$ total cost which is impractical for even the short toy data examples provided in the paper.
-To get out of this re-introduces the problem we set out to solve: how to choose a range of possible durations over which to sample.
-One way to do this in a manner that is guaranteed not to break, is to allow any duration up to the length of the sequence.
+This requires $O(T^3K^2)$ total cost which is impractical for even the short toy data examples provided in the paper.
+To escape this cost re-introduces the problem we set out to solve: how to choose a range of possible durations over which to sample.
+One way to do this, in a manner that is guaranteed not to break, is to allow any duration up to the length of the sequence.
 Since this is computationally infeasable (see the caption to Figure 4), we can, instead, show the achieved expected joint-likelihood of the model and the data as we increase $d_\mathrm{max}$.
-This unsurprisingly breaks exactly as we would expect it to.
-This graph shows the mean likelihood of the samples after burn-in as the maximum duration increases, and shows one way that the truncated forward/backward algorithm suffers when the true durations are not in the support of the truncation windows.
-We have placed the code for this experiment in the code repository and include a quick graph below.
-The mean likelihood of the beam sampler is shown in purple for comparison.
 
-MIKE: IF THERE ARE NO STATES OF DURATIONS GREATER THAN 25 IN THE DATA THEN I DONT BELIEVE THIS FIGURE.
-IF $d_{max}$ IS SET LARGER THAN THE BIGGEST OBSERVED SEQUENCE THEN THESE FIGURES SHOULD OVERLAP.
-I DON'T KNOW WHAT DATA THIS FIGURE CORRESPONDS TO AND WHAT DURATIONS ARE IN THE DATA.
-IT MIGHT MAKE SENSE TO RUN THESE EXPERIMENTS OUT TO OVERLAP.
-IT SHOULD BE POSSIBLE.
+This unsurprisingly performs poorly exactly as we would expect it to: until $d_\mathrm{max}$ is increased to a value that captures the maximum duration the model fails to adequately capture the true dynamics.
+Figure \ref{fig} shows the likelihood of 5 samples from the forward backward algorithm as the maximum duration increases, and shows one way that the truncated forward/backward algorithm suffers when the true durations are not in the support of the truncation windows.
+For comparison, the likelihood of five state sequences sampled using the full beam sampler is shown in red, and the likelihood of the truth is shown in green.
+While we would enjoy to place this figure in the manuscript, we unfortunately do not have space.
+Instead, we have placed the code for this experiment in the code repository referenced in the paper.
 
 \begin{figure}
-	\includegraphics[width=0.8\textwidth]{../pic/likelihood_over_dmax.pdf}
-	\caption{Mean likelihood over samples after burn-in as the range of possible durations increases (blue line). Shown for comparison is the mean likelihood of the beam sampler (purple line)}
+	\includegraphics[width=\textwidth]{../pic/likelihood_over_dmax.pdf}
+	\caption{Likelihood of sampled state sequences from the forward backward algorithm as the range of possible durations increases (blue lines).
+Shown for comparison is the likelihood of the true state sequence (green line), and the likelihood of samples from the beam sampler (red lines).}
+	\label{fig}
 \end{figure}
 
-\section*{Reviwewr 2}
+\section*{Reviewer 2}
 
-We thank the reviewer for their comments, we have addressed the typos.
+We thank the reviewer for their comments; we have addressed the typos.
+The parameter estimation suffers in the second example as we're not able to completely separate the states that have identical observation distributions.
+The observation distributions all have identical variances, so the separation between states is not due to the observation variance.
+We do not have an explicit mechanism for controlling state switching in the current algorithm implementation; this would be interesting future work.
+
+Having said this, we are able to show that the algorithm can identify separate states in the second experiment and perform reasonable inference.
+Examining this specific problem of states that are only seperable by duration would also be interesting future work.
+It's especially interesting, for example, when studying behaviour in animals (an early motivation for the current paper), as many behaviours are distinguishable by eye only because they spend a different amount of time performing each behaviour: consider `pausing', vs `stunned' vs `dead' - in a video of fruit flies all these behaviours look the same (i.e.
+have the same observation distribution) over a few frames yet have large duration distribution differences.
 
@@
 Reviewer 3 has highlighted two omissions in the paper.
-The first was that we did not make clear the fact that the Explicit Duration HMM is a type of Hidden Semi Markov Model, which is now made clear in the second paragraph of the introduction, along with an additional citation to an HSMM review detailing the relationship.
+The first was that we did not make clear the fact that the Explicit Duration HMM is a type of Hidden Semi Markov Model (HSMM), which is now made clear in the second paragraph of the introduction, along with an additional citation to a review of HSMMs detailing the relationship.
 The second was that we did not report on the results of the transition rate parameter estimates.
@@
 
+Reviewer 3 also suggested a comparison with different model structures.
+We hope that the clarification of the EDHMM as a type of HSMM addresses this point somewhat, and that clarifies our aim of presenting a novel inference method in a standard model form.
+Space limitations restrict discussion of alternative model forms; we hope that the included reference from Yu et al provides ample review.
+
 \section*{Reviewer 4}
 
-We thank the reviewer for their comments, and have fixed the typos. 
+We thank the reviewer for their comments, and have fixed the specified typos.
+
 \end{document}

### 2012-01-14 tex/abstract.tex
final edits

@@
-Explicit-state-duration hidden Markov models (EDHMM) are HMMs that have latent states consisting of both discrete state-indicator and discrete state-duration random variables.
+In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free, black-box inference procedure for Explicit-state-duration hidden Markov models (EDHMM).
+EDHMMs are HMMs that have latent states consisting of both discrete state-indicator and discrete state-duration random variables.
 In contrast to the implicit geometric state duration distribution possessed by the standard HMM, EDHMMs allow the direct parameterisation and estimation of per-state duration distributions.
 As most duration distributions are defined over the positive integers, truncation or other approximations are usually required to perform EDHMM inference.
-In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free, black-box inference procedure for EDHMMs.
-%We illustrate the performance gains of this approach using synthetic data.

### 2012-01-14 tex/model.tex
small edits, mostly to do with computation complexity. One graph left...
final edits

@@
 We define the observation sequence $\mathcal{Y} = \{y_1, y_2, \ldots, y_T\}$; the latent state sequence $\mathcal{X} = \{x_0, x_1, x_2, \ldots, x_T\}$; and the remaining time in each segment $\mathcal{D} = \{ d_1, d_2, \ldots, d_T\}$, where $x_t \in \{ 1, 2, \ldots, K\}$ with $K$ the maximum number of states, $d_t \in \{1, 2, \ldots \}$, and $y_t \in \mathbb{R}^n$.
-We assume that the Markov chain on the latent states is homogenous, i.e., that $p(x_t = i | x_{t-1}=j, A) = a_{i,j} \forall t$ where $A$ is a $K\times K$ matrix with element $a_{i,j}$ at row $i$ and column $j.$  The prior on $A$ is row-wise Dirichlet with zero prior mass on self-transitions, i.e.
-$p(a_{i,:}) = \mathrm{Dir}({1}/{K-1}, \ldots, 0 , \ldots {1}/{K-1})$ where $a_{i,:}$ is a row vector and the $i$th Dirichlet parameter is $0.$  Each state is imbued with its own duration distribution $p(d_t | x_t = k) = p(d_t | \lambda_{k})$ with parameter $\lambda_k$.
+We assume that the Markov chain on the latent states is homogenous, i.e., that $p(x_t = j | x_{t-1}=i, A) = a_{i,j} \forall t$ where $A$ is a $K\times K$ matrix with element $a_{i,j}$ at row $i$ and column $j.$  The prior on $A$ is row-wise Dirichlet with zero prior mass on self-transitions, i.e.
+$p(a_{i,:}) = \mathrm{Dir}({1}/{(K-1)}, \ldots, 0 , \ldots {1}/{K-1})$ where $a_{i,:}$ is a row vector and the $i$th Dirichlet parameter is $0.$  Each state is imbued with its own duration distribution $p(d_t | x_t = k) = p(d_t | \lambda_{k})$ with parameter $\lambda_k$.
 Each duration distribution parameter is drawn from a prior $p(\lambda_k)$ which can be chosen in an application specific way.
@@
 Each state is also imbued with an observation generating distribution $p(y_t  | x_t = k) = p(y_t | \theta_{k})$ with parameter $\theta_k$.
-Each observation distribution parameter is
-%given 
-drawn from a prior $p(\theta_k)$ also to be chosen according to the application.
+Each observation distribution parameter is drawn from a prior $p(\theta_k)$ also to be chosen according to the application.
 The set of all observation distribution parameters is $\theta.$  In the following exposition, explicit conditional dependencies on component distribution parameters are omitted to focus on the particulars unique to the EDHMM.
@@
     \label{eq:joint}
-    p(\mathcal{X},\mathcal{D},\mathcal{Y}) =  \\ 
-    p(x_0)p(d_0)\prod_{t=1}^T p(y_t | x_t, \theta) p(x_t | x_{t-1}, d_{t-1}, A) p(d_t | x_{t}, d_{t-1}, \lambda)
+    p(\mathcal{X},\mathcal{D},\mathcal{Y}) =  
+    p(x_0)p(d_0) \\ \prod_{t=1}^T p(y_t | x_t, \theta) p(x_t | x_{t-1}, d_{t-1}, A) p(d_t | x_{t}, d_{t-1}, \lambda)
 \end{multline}
